### 2023-01-20 paper/localpsf.tex
work on the ice section

The true parameter field, $\basalfriction_\text{true}$, and corresponding velocity field, $\velocity_\text{true}$, are shown in Figure \ref{fig:true_beta_u}. We generate synthetic observations by restricting the tangential component of $\velocity_\text{true}$ to the top surface to get
\begin{equation*}
\observations_\text{true} = \tangentop \velocity_\text{true} |_{\Gamma_t}.
\end{equation*}
We add $1\%$ multiplicative Gaussian noise to all components of the finite element vector, $\bm{\observations}_\text{true}$, corresponding to the function $\observations_\text{true}$, which yields noisy synthetic observations, $\observations_\text{obs}$ which are used for the inversion. That is, 
\begin{equation*}
\left(\bm{\observations}_\text{obs}\right)_j := \left(\bm{\observations}_\text{true}\right)_j + 0.01 \cdot \boldsymbol{\xi}_j \cdot \left|\left(\bm{\observations}_\text{true}\right)_j\right|,
\end{equation*}
where $j$ ranges over all entries of the vector of coefficients representing the finite element functions $\observations_\text{true}$ and $\observations_\text{obs}$, and $\boldsymbol{\xi}_j$ are independent and identically distributed (i.i.d) random numbers drawn from the standard normal distribution. We also define the noise function, $\xi$, to be the finite element function with $j$th vector entry $\boldsymbol{\xi}_j$.

The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
\begin{equation}
	\label{eq:deterministic_optimization_problem}
	\min_{\basalfriction} \quad J(\basalfriction) := J_d(\basalfriction) + J_r(\basalfriction).
\end{equation}
The first term in the objective function is
\begin{equation*}
	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS.
\end{equation*}
We call this term the data misfit, because it measures the difference between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\velocity = \velocity(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
\begin{equation*}
	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{K}(\basalfriction - \basalfriction_0)\right|^2\,dx,
\end{equation*}
is a Bilaplacian regularization term. Specifically, $\basalfriction_0$ is the constant function $\basalfriction_0(x)=10.5$, and $\mathcal{K}$ is the inverse of the solution operator for the following elliptic PDE:
\begin{align*}
-\gamma\Delta u +\delta\,u&=\regrhs\,\,\,\,\text{ in }\Omega,\\
\gamma \nabla u\cdot\normalvec+\regrobincoeff\,u&=0\,\,\,\text{ on }\partial\Omega.
\end{align*}
Recall that $\Omega$ is the 2D flattened version of the basal surface $\Gamma_b$. Here $\regrhs$ is a generic forcing term, and $\regrobincoeff=\sqrt{\delta \gamma}/1.42$ is a Robin boundary condition coefficient \cite{Roininen14}, and these quantities are different from $\bodyforce$ and $\stokesrobincoeff$ in the Stokes equation, \eqref{Stokeseqn}.
In all results except one, $\gamma$ is chosen so that the Morozov discrepancy principle is satisfied, i.e.,
\begin{equation*}
	\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS = \int_{\Gamma_{t}}||\xi||_{2}^{2}\,dS.
\end{equation*}
The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness of our preconditioner. The constant $\delta$ is chosen so that the correlation length of functions drawn from the normal distribution with covariance $\mathcal{K}^{-2}$, given by $L=\sqrt{\gamma/\delta}$, is $1/10$th the radius of the domain. 
%, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.

We solve the inverse problem with an inexact Newton preconditioned conjugate gradient (PCG) scheme \cite[Section 7.1]{NocedalWright99}, using $\basalfriction_0$ as the initial guess. The discretized parameter, $\bm{\basalfriction}$ is updated iteratively according to the formula
$
	\bm{\basalfriction}_{k+1} = \bm{\basalfriction}_k + \alpha_k \bm{\searchdir}_k,
$
where $\bm{\searchdir}_k$ approximately solves one of the following Newton linear systems,
\begin{equation}
	\label{eq:newton_system}
	\bm{H} \bm{\searchdir}_k = - \bm{g}_k \qquad \text{or} \qquad \bm{H}_\text{gn} \bm{\searchdir}_k = - \bm{g}_k,
\end{equation}
and the step length $\alpha_k \in \mathbb{R}$ is chosen by line search to satisfy the Armijo conditions \cite[Section 3.1]{NocedalWright99}. Here $\bm{g}_k$ is the discretized gradient of $J$, $\bm{H}$ is the discretized Hessian, and $\bm{H}_\text{gn}$ is the discretized Gauss-Newton Hessian, and these derivatives are evaluated at $\bm{\basalfriction}_k$. 
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
We use $\bm{H}_\text{gn}$ for the first five iterations, and $\bm{H}$ for all subsequent iterations.  The Newton iteration terminates when $||\bm{g}_k||/||\bm{g}_0|| < 10^{-8}$. Systems \eqref{eq:newton_system} are solved inexactly using an inner PCG iteration, which terminates when the norm of the residual for the linear system is less than $\min \left(0.5, \sqrt{||\bm{g}_k||}\right)||\bm{g}_k||$. PCG only requires applying $\bm{H}$ (or $\bm{H}_\text{gn}$) to vectors. Each application of $\bm{H}$ (or $\bm{H}_\text{gn}$) to a vector requires solving one incremental forward and one incremental adjoint Stokes PDE of the form \eqref{Stokeseqn}, but with different right hand sides.

\begin{figure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m1_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m3_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m4_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m6_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{The log basal friction parameter after $1$ (left), $3$ (second from left), and $4$ (third from left) Newton iterations, and the final Newton iterate (right). The PSF (5) preconditioner is constructed between Newton iterations 3 and 4.}
	\label{fig:newton_iterates} 
\end{figure}

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$. Straightforward calculation shows that the discretized Hessian of the regularization term is given by
$
(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
$
where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques. We factor the overall Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$, using fast H-matrix methods, then use the factorization as a preconditioner.

>>>>

The observations are generated by adding $1\%$ multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. The true basal sliding
coefficient field and the true velocity field , which is obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

To reconstruct the basal sliding parameter field $\basalfriction$, we
formulate an inverse problem governed by the Stokes
problem~\eqref{Stokeseqn}. In particular this problem is formulated as
a nonlinear least squares optimization problem, whose cost function
consists of a misfit term (between the observations and model output)
and a bi-Laplacian regularization term
following~\cite{IsaacPetraStadlerEtAl15}. The regularization is
centered at the constant function $q_0(x) = 10.5$. To mitigate
boundary effects we use a constant coefficient Robin boundary
condition as in~\cite{Roininen14}. The parameters are chosen so that
the Green's function of the Hessian of the regularization has a
characteristic length of $0.25$ of the radius of the domain. For the
specific setup, we refer the reader
to~[sec.22]\cite{VillaPetraGhattas21}. We choose the regularization
parameter, $\gamma$, that controls the overall strength of the
regularization using Morozov's discrepancy principle. The optimal
value of $\gamma$ is $7.3 \times 10^3$, which is used for all the
numerical results except for Figure~\ref{fig:gamma_sweep}, where we
vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness
of the proposed preconditioner.

%% from here we summarize --------
%The true parameter field, $\basalfriction_\text{true}$, and corresponding velocity field, $\velocity_\text{true}$, are shown in Figure \ref{fig:true_beta_u}. We generate synthetic observations by restricting the tangential component of $\velocity_\text{true}$ to the top surface to get
%\begin{equation*}
%\observations_\text{true} = \tangentop \velocity_\text{true} |_{\Gamma_t}.
%\end{equation*}
%We add $1\%$ multiplicative Gaussian noise to all components of the finite element vector, $\bm{\observations}_\text{true}$, corresponding to the function $\observations_\text{true}$, which yields noisy synthetic observations, $\observations_\text{obs}$ which are used for the inversion. That is, 
%\begin{equation*}
%\left(\bm{\observations}_\text{obs}\right)_j := \left(\bm{\observations}_\text{true}\right)_j + 0.01 \cdot \boldsymbol{\xi}_j \cdot \left|\left(\bm{\observations}_\text{true}\right)_j\right|,
%\end{equation*}
%where $j$ ranges over all entries of the vector of coefficients representing the finite element functions $\observations_\text{true}$ and $\observations_\text{obs}$, and $\boldsymbol{\xi}_j$ are independent and identically distributed (i.i.d) random numbers drawn from the standard normal distribution. We also define the noise function, $\xi$, to be the finite element function with $j$th vector entry $\boldsymbol{\xi}_j$.
%
%The reconstructed parameter field $\basalfriction$ is found as the solution to an optimization problem of the following form:
%\begin{equation}
%	\label{eq:deterministic_optimization_problem}
%	\min_{\basalfriction} \quad J(\basalfriction) := J_d(\basalfriction) + J_r(\basalfriction).
%\end{equation}
%The first term in the objective function is
%\begin{equation*}
%	J_d(\basalfriction) := \frac{1}{2}\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS.
%\end{equation*}
%We call this term the data misfit, because it measures the difference between the observed data and the predicted data based on a candidate parameter $\basalfriction$. Here $\velocity = \velocity(\basalfriction)$ denotes the velocity field solving $\ref{Stokeseqn}$ for the given parameter $\basalfriction$. The second term,
%\begin{equation*}
%	J_r(\basalfriction) := \frac{1}{2}\int_{\Omega}\left| \mathcal{K}(\basalfriction - \basalfriction_0)\right|^2\,dx,
%\end{equation*}
%is a Bilaplacian regularization term. Specifically, $\basalfriction_0$ is the constant function $\basalfriction_0(x)=10.5$, and $\mathcal{K}$ is the inverse of the solution operator for the following elliptic PDE:
%\begin{align*}
%-\gamma\Delta u +\delta\,u&=\regrhs\,\,\,\,\text{ in }\Omega,\\
%\gamma \nabla u\cdot\normalvec+\regrobincoeff\,u&=0\,\,\,\text{ on }\partial\Omega.
%\end{align*}
%Recall that $\Omega$ is the 2D flattened version of the basal surface $\Gamma_b$. Here $\regrhs$ is a generic forcing term, and $\regrobincoeff=\sqrt{\delta \gamma}/1.42$ is a Robin boundary condition coefficient \cite{Roininen14}, and these quantities are different from $\bodyforce$ and $\stokesrobincoeff$ in the Stokes equation, \eqref{Stokeseqn}.
%In all results except one, $\gamma$ is chosen so that the Morozov discrepancy principle is satisfied, i.e.,
%\begin{equation*}
%	\int_{\Gamma_{t}}||\observations_\text{obs} - \tangentop\velocity||_{2}^{2}\,dS = \int_{\Gamma_{t}}||\xi||_{2}^{2}\,dS.
%\end{equation*}
%The value of $\gamma$ satisfying the Morozov discrepancy principle is $\gamma=7.3 \times 10^3$. The exception is Figure \ref{fig:gamma_sweep}, where we vary $\gamma$ to study how changing $\gamma$ impacts the effectiveness of our preconditioner. The constant $\delta$ is chosen so that the correlation length of functions drawn from the normal distribution with covariance $\mathcal{K}^{-2}$, given by $L=\sqrt{\gamma/\delta}$, is $1/10$th the radius of the domain. 
%%, which is halfway between the maximum and minimum values of $\basalfriction_\text{true}$.
%

We solve the inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and Armijo line search for
globalization~\cite{NocedalWright99}. The (discrete) Newton updates,
$\bm{\searchdir}_k$, are obtained by solving 
\begin{equation}
	\label{eq:newton_system}
	\bm{H} \bm{\searchdir}_k = - \bm{g}_k \qquad \text{or} \qquad \bm{H}_\text{gn} \bm{\searchdir}_k = - \bm{g}_k,
\end{equation}
The initial guess is chosen as the discretization of the constant
function $q_0$. Here $\bm{g}_k$, $\bm{H}$ and $\bm{H}_\text{gn}$ are
the discretized gradient Hessian and Gauss-Newton Hessian of the
inverse problem cost function, respectively.
%The Gauss-Newton Hessian is Hessian of $J$, except in the data misfit, $J_d$, the nonlinear mapping $\basalfriction \mapsto \tangentop \velocity$ is replaced by its local linearization at $\basalfriction_k$.
To ensure positive definiteness of the Hessian we use
$\bm{H}_\text{gn}$ for the first five iterations, and $\bm{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$||\bm{g}_k||/||\bm{g}_0|| < 10^{-8}$. Systems
\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which terminates when the norm of the residual for the
linear system is less than $\min \left(0.5,
\sqrt{||\bm{g}_k||}\right)||\bm{g}_k||$. %PCG only requires applying
%$\bm{H}$ (or $\bm{H}_\text{gn}$) to vectors. Each application of
%$\bm{H}$ (or $\bm{H}_\text{gn}$) to a vector requires solving one
%incremental forward and one incremental adjoint Stokes PDE of the form
%\eqref{Stokeseqn}, but with different right hand sides.
\begin{figure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m1_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m3_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m4_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.1]{m6_psf.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{The log basal friction parameter after $1$ (left), $3$ (second from left), and $4$ (third from left) Newton iterations, and the final Newton iterate (right). The PSF (5) preconditioner is constructed between Newton iterations 3 and 4.}
	\label{fig:newton_iterates} 
\end{figure}

We use the framework described in this paper to generate Hessian preconditioners. We build H-matrix approximations of the data misfit Gauss-Newton Hessian (the term in $\bm{H}_\text{gn}$ that arises from the data misfit) using 1, 5, and 25 batches. These approximations are denoted by PSF (1), PSF (5), and PSF (25), respectively. The resulting H-matrices, which may be indefinite due to approximation error, are modified to be symmetric positive definite via the procedure described in Appendix \ref{sec:make_spd}, with $l=2$. We approximate $\bm{H}_\text{gn}$ rather than $\bm{H}$ because $\bm{H}$ more often has negative values in its integral kernel, especially when one is away from the optimal point. Our results show that our preconditioners are also good preconditioners for $\bm{H}$. The H-matrix approximation of the data misfit Gauss-Newton Hessian is added to an H-matrix approximation of the Hessian of the regularization term to form an H-matrix approximation of $\bm{H}_\text{gn}$.
%
%Straightforward calculation shows that the discretized Hessian of the regularization term is given by
%$
%(\gamma \bm{K} + \delta \bm{M}) \bm{M}^{-1} (\gamma \bm{K} + \delta \bm{M})
%$
%where $\bm{K}$ is a finite element stiffness matrix with appropriate boundary conditions, and $\bm{M}$ is the finite element mass matrix. These matrices are sparse, so we form H-matrix representations of these matrices and combine them using standard sparse H-matrix techniques.
%
We note that the Hessian of the regularization is a combination of
stiffness and mass matrices, which are sparse. Therefore, we form
H-matrix representations of these matrices and combine them using
standard sparse H-matrix techniques.  We factor the overall
Gauss-Newton Hessian approximation, denoted $\bm{\preconditioner}$,
using fast H-matrix methods, then use the factorization as a
preconditioner.
